Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-21 _tex/index.tex
Built site for gh-pages

However, since the discriminal process of a single stimulus is not
directly observable, the \emph{law of comparative judgment} becomes
essential. This law states that in pairwise comparisons, the stimulus
positioned further along the trait continuum is perceived as having a
higher level of that trait. Thus, the theory assumes the observed
dichotomous outcome is determined by the distribution of the difference
between the underlying discriminal processes of the stimuli, referred to
as the \emph{discriminal difference}. This suggests that the outcome
depends on the relative distance between stimuli, rather than their
absolute positions on the trait continuum.

These concepts can be illustrated through an example. For instance, in
the context of evaluating text quality,
Figure~\ref{fig-discriminal_process} could depict the underlying
discriminal process distributions for two written texts, highlighting
differences in their discriminal dispersions and modal discriminal
processes along the quality trait continuum. Furthermore,
Figure~\ref{fig-discriminal_difference} could display the discriminal
difference distribution for these texts, showing that text A is
perceived to exhibit significantly higher quality than text B, as
indicated by the shaded gray area. Consequently, the dichotomous outcome
of this comparison would favor text A.

@@

The assumption of \emph{zero correlation between stimuli} is more easily
understood through an example. For instance, when using pairwise
comparisons to evaluate text quality, the assumption suggests that a
judge's perception of the trait in one text does not influence his
perception of the same trait on another. Thurstone attributed this
independence to the cancellation of potential judges' biases, driven by
two opposing and equally weighted effects occurring during the pairwise
comparisons \citep[pp.~268]{Thurstone_1927b}. This cancellation was
mathematically demonstrated by \citet{Andrich_1978}, assuming
discriminal processes with additive biases. However, it is easy to
imagine two scenarios where the zero correlation assumption almost
certainly does not hold: when the pairwise comparison involves
multidimensional, complex traits with heterogeneous stimuli, and when an
additional hierarchical structure is relevant to the stimuli.

In the first scenario, evidence suggests that when judges are confronted
with multi-dimensional, complex traits and heterogeneous stimuli, such
as when evaluating written text quality, their assessments are
influenced by multiple, intricate aspects of the stimuli
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. In
such cases, it is not inconceivable that these various aspects of the
stimuli unevenly influence the judges' perception of one text over
another, leading to biases that resist cancellation, as the effects of
the factors involved in the comparison are neither equally weighted nor
opposing. For example, this may occur when a judge, while assessing the
argumentative quality of a text, places disproportionate emphasis on
grammatical errors, ultimately favoring texts with fewer errors but
weaker arguments. Although the literature does not provide direct
evidence for this specific argument, studies such as
\citet{Pollitt_et_al_2003} do offer evidence of judges' biases,
supporting the idea that the factors involved in pairwise comparisons
may not always cancel out.

@@

However, the statistical literature advises caution when using estimated
scores to conduct additional analyses and tests while warning against
eliminating data through ad-hoc univariate procedures. A key
consideration is that BTL scores are parameter estimates that inherently
carry uncertainty. Ignoring this uncertainty can bias the analysis and
reduce the precision of hypothesis tests. Notably, the direction and
magnitude of such biases are often unpredictable. Results may be
attenuated, exaggerated, or remain unaffected depending on the degree of
uncertainty in the scores and the actual effects being tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Moreover, eliminating data using ad-hoc univariate procedures using
these scores compounds this issue by excluding possible valuable
information from the analyses \citep{McElreath_2020}, further
introducing bias. Finally, the reduced precision in hypothesis tests
weakens their statistical power, increasing the likelihood of type-I or
type-II errors \citep{McElreath_2020}.

@@

\subsection{Assessment design features and their role on reliability and
validity}\label{sec-theory-issue3}

>>>>

However, since the discriminal process of a single stimulus is not
directly observable, the \emph{law of comparative judgment} becomes
essential. This law states that in pairwise comparisons, the stimulus
positioned further along the trait continuum is perceived as having a
higher level of that trait. Thus, the theory assumes the observed
dichotomous outcome is determined by the distribution of the difference
between the underlying discriminal processes of the stimuli, referred to
as the \emph{discriminal difference}. This indicates that the outcome
depends on the relative distance between stimuli, rather than their
absolute positions on the trait continuum.

These concepts are more easily understood through an example. For
instance, in the context of evaluating text quality,
Figure~\ref{fig-discriminal_process} could depict the underlying
discriminal process distributions for two written texts, highlighting
differences in their discriminal dispersions and modal discriminal
processes along the quality trait continuum. Furthermore,
Figure~\ref{fig-discriminal_difference} could display the discriminal
difference distribution for these texts, showing that text A is
perceived to exhibit significantly higher quality than text B, as
indicated by the shaded gray area. Consequently, the dichotomous outcome
of this comparison would favor text A.

@@

The assumption of \emph{zero correlation between stimuli} is more easily
understood through an example. For instance, when using pairwise
comparisons to evaluate text quality, the assumption suggests that the
judge's perception of the trait in one text does not influence his
perception of the same trait on another. Thurstone attributed this
independence to the cancellation of potential judges' biases, driven by
two opposing and equally weighted effects occurring during the pairwise
comparisons \citep[pp.~268]{Thurstone_1927b}. This cancellation was
mathematically demonstrated by \citet{Andrich_1978}, assuming
discriminal processes with additive biases. However, it is easy to
imagine two scenarios where the zero correlation assumption almost
certainly does not hold: when the pairwise comparison involves
multidimensional, complex traits with heterogeneous stimuli, and when an
additional hierarchical structure is relevant to the stimuli.

In the first scenario, evidence suggests that when judges are confronted
with multi-dimensional, complex traits and heterogeneous stimuli, such
as when evaluating written text quality, their assessments are
influenced by multiple, intricate aspects of the stimuli
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. In
such cases, it is not inconceivable that various aspects of the stimuli
unevenly influence the judges' perception of one text over another,
leading to biases that resist cancellation, as the effects of the
factors involved in the comparison are neither equally weighted nor
opposing. For example, this may occur when a judge, while assessing the
argumentative quality of a text, places disproportionate emphasis on
grammatical errors, ultimately favoring texts with fewer errors but
weaker arguments. Although the literature does not provide direct
evidence for this specific argument, studies such as
\citet{Pollitt_et_al_2003} do offer evidence of judges' biases,
supporting the idea that the factors involved in the pairwise
comparisons may not always cancel out.

In the second scenario, although the CJ literature has acknowledged the
existence of additional hierarchical structures relevant to the stimuli,
the statistical handling of this extra source of correlation has been
inadequate. For example, in cases where data included multiple samples
of stimuli within different individuals, researchers have often relied
on using (averaged) estimated BTL scores to conduct further analyses and
tests at the hierarchical level of the individuals
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach has the notable limitation of ignoring the
uncertainty associated with the scores (refer to section
Section~\ref{sec-theory-issue2} for a detailed discussion of this
issue).

In contrast, the assumption of \emph{equal dispersion between stimuli}
suggests that the perceived trait of all texts varies around their modal
discriminal process in the same way across comparisons. While Thurstone
noted that the assumption of equal dispersion may be violated when
``dealing with less conspicuous attributes or with less homogeneous
stimuli'' \citep[pp.~374]{Thurstone_1927a}, no study explicitly proposes
that this assumption may also be violated due to the presence of an
additional hierarchical structure relevant to the texts. One such
scenario could occur when comparing university and secondary school
students. In this case, university students may consistently (or more
precisely) produce higher-quality texts, while secondary school students
might exhibit a broader range of writing abilities, leading to greater
variability in their text quality. Despite the contrived nature of the
example, it helps to effectively illustrate how assuming equal
dispersions in the texts can overlook meaningful differences in the
reliability of text quality across groups or individuals.

@@

However, the statistical literature advises caution when using estimated
scores for additional analyses and tests, as well as when eliminating
data through ad hoc univariate procedures. A key consideration is that
BTL scores are parameter estimates that inherently carry uncertainty.
Ignoring this uncertainty can bias the analysis and reduce the precision
of hypothesis tests. Notably, the direction and magnitude of such biases
are often unpredictable. Results may be attenuated, exaggerated, or
remain unaffected depending on the degree of uncertainty in the scores
and the actual effects being tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Moreover, excluding data using ad hoc univariate procedures can compound
these issues by discarding potentially valuable information, further
exacerbating the bias \citep{Zimmerman_1994, McElreath_2020}. Finally,
the reduced precision in hypothesis tests diminishes their statistical
power, increasing the likelihood of type-I or type-II errors
\citep{McElreath_2020}.

@@

\subsection{The assessment design features and their role on reliability
and validity}\label{sec-theory-issue3}
